---
title: "Week2"
author: "Lisa Wahlen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: import Data

Create a new R- (or RMarkdown/Quarto-) file and begin with the following lines of code (adjust the path to your csv file accordingly).

### load the libraries

```{r}
library(readr) # to import tabular data (e.g. csv)
library(dplyr) # to manipulate (tabular) data
library(ggplot2) # to visualize data
library(sf) # to handle spatial vector data
library(terra) # To handle raster data
library(lubridate) # To handle dates and times
```

### Import the downloaded csv

```{r}
wildschwein_BE <- read_delim("data/wildschwein_BE_2056.csv", ",")

wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), crs = 2056, remove = FALSE)
```

Note:

-   that this dataset is already converted to EPSG 2056

-   the coordinates are stored in the columns (`E`/`N`)

-   setting `remove = FALSE` preserves the original (`E`/`N`) columns, which come in handy later on
